Appendix 4: species distribution model comparison between inla and sdmTMB

### 2021-12-16 doc/appendix-binomial-inla.Rmd
a little more refining of inla example

```{r set-priors, echo = TRUE, cache=FALSE}
range_min <- 5
range_prob <- 0.05
sigma_max <- 5
sigma_prob <- 0.05
```

@@

```{r inla-mesh, echo = TRUE, cache=FALSE}
loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)

spde <- inla.spde2.pcmatern(
  mesh = inla_mesh, # an smdTMB mesh can be inserted here with sdm_mesh$mesh
    prior.range = c(range_min, range_prob),
    prior.sigma = c(sigma_max, sigma_prob)) 
```

```{r plot-inla-mesh, fig.asp=0.9, out.height="3in"}
plot(sdm_mesh);title(main = "sdmTMB mesh");plot(spde$mesh)
```

@@

m <- sdmTMB(present ~ 0 + as.factor(year), 
  data = d, 
  mesh = sdm_mesh, 
  time = "year",
  spatial = "off", 
  spatiotemporal = "AR1", 
  family = binomial(link = "logit"),
  priors = sdmTMBpriors(
    matern_st = pc_matern(
      range_gt = range_min, range_prob = range_prob, #P(range > 5) = 1-0.05
      sigma_lt = sigma_max, sigma_prob = sigma_prob #P(sigma < 5) = 1-0.05
    ),
    b = normal(0, 100)
  ))

@@

```{r compare-plot, out.width="6in"}
ggplot(post_inla2) + 
  geom_histogram(data=post_sdmTMB2, aes(.value, after_stat(density)), bins = 30, 
                 fill = "blue", alpha = 0.2) + 
  geom_line(data=post_inla2, aes(x, y), inherit.aes = F) + 
  facet_wrap(~variable, scales = "free") 
```
# Distribution map

We can predict on a the grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to produce a map of occurrence probabilities in each year.

```{r sdm-predictions}
nd <- filter(qcs_grid, year >= 2011)
pred <- predict(m, newdata = nd)

plot_map <- function(dat, column) {
ggplot(dat, aes_string("X", "Y", fill = column)) +
    geom_raster() +
    facet_wrap(~year) +
    coord_fixed()
}
```

```{r maps, cache=FALSE, out.width="6in"}
plot_map(pred, "plogis(est)") +
  scale_fill_viridis_c() +
  ggtitle("Prediction (fixed effects + all random effects)")
```
Similar predictions are more complicated to extract from INLA, but can be plotted for the original sample locations.

```{r inla-pred, out.width="6in"}
idx <- inla.stack.index(sdat, 'stdata')$data

d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]

ggplot(d) + geom_point(aes(X, Y, colour = inla_pred), size = 2) + facet_wrap(~year) + 
  scale_colour_viridis_c() +
  coord_fixed() +
  ggtitle("Prediction (fixed effects + all random effects)")
```

>>>>

```{r set-priors, echo = TRUE, cache=FALSE}
range_min <- 5
sigma_max <- 5
prior_prob <- 0.05
```

@@

```{r inla-mesh, echo = TRUE, cache=FALSE}
loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)
spde <- inla.spde2.pcmatern(
  mesh = inla_mesh, # an smdTMB mesh can be inserted here with sdm_mesh$mesh
    prior.range = c(range_min, prior_prob),
    prior.sigma = c(sigma_max, prior_prob)) 
```

```{r plot-inla-mesh, fig.asp=0.9, out.height="2.5in"}
plot(sdm_mesh);title(main = "sdmTMB mesh");plot(spde$mesh)
```

@@

m <- sdmTMB(present ~ 0 + as.factor(year), 
  data = d, 
  mesh = sdm_mesh, 
  time = "year",
  spatial = "off", 
  spatiotemporal = "AR1", 
  family = binomial(link = "logit"),
  priors = sdmTMBpriors(
    matern_st = pc_matern(
      range_gt = range_min, range_prob = prior_prob, #P(range > 5) = 1-0.05
      sigma_lt = sigma_max, sigma_prob = prior_prob #P(sigma < 5) = 1-0.05
    ),
    b = normal(0, 100)
  ))

@@

```{r compare-plot, out.width="6in"}
ggplot(post_inla2) + 
  geom_histogram(data=post_sdmTMB2, aes(.value, after_stat(density)), bins = 30, 
                 fill = "blue", alpha = 0.2) + 
  geom_line(data=post_inla2, aes(x, y), inherit.aes = F) + 
  facet_wrap(~variable, scales = "free") 
```

# Annual distribution maps

We can predict on a the grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year.

```{r sdm-predictions}
nd <- filter(qcs_grid, year >= 2011)
pred <- predict(m, newdata = nd) 
```

```{r maps, cache=FALSE, out.width="4in"}
ggplot(pred) + geom_raster(aes(X, Y, fill = plogis(est))) + 
  facet_wrap(~year) +   scale_fill_viridis_c() + coord_fixed() 
```

Similar predictions are more complicated to extract from INLA, but can be plotted for the original sample locations.

```{r inla-pred, out.width="4in"}
idx <- inla.stack.index(sdat, 'stdata')$data
d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
ggplot(d) + geom_point(aes(X, Y, colour = inla_pred), size = 2) + 
  facet_wrap(~year) + scale_colour_viridis_c() + coord_fixed() 
```
